**OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units**

### 2017-02-02 submission/response_to_reviewers.md

---
title: ""
output:
  pdf_document:
    includes:
      in_header: header.tex
fontsize: 11pt
geometry: margin=1.0in
---

## Reviewer 1

### Major comments
**Optimizing a criterion is a good idea. It would be useful to clarify if each of the other algorithms aims to optimize some other criterion and if so which one. There is some text about this in the discussion, but this information might work better earlier in the manuscript. It would also be helpful to see how each algorithm performs on these other metrics (e.g., minimizing FPs). You can argue that MCC is a better criterion to use, but it is also good to show if the methods that aim to optimize another criterion actually do so and how MCC-based OptiClust does with regards to these metrics. In addition, some further discussion about the relative costs of FPs and FNs in typical metagenomics applications and downstream studies would provide some context for your argument to use MCC.**

We have added text to the Introduction to describe that the furthest and nearest neighbor algorithms seek to minimize the number of FPs and FNs, respectively (***LXXX***). We have also added text to the Discussion (***LXXX***) pointing out that More FPs would increase the number of OTUs while more FNs would collapse OTUs together. It is difficult to know which is worse since both alpha and beta diversity metrics are tied to the number and composition of OTUs. As we point out, this is why it is important to jointly minimize the number of FPs and FNs.

**For the simulations, it would be nice to show the true values of some of the output measures (e.g., number of clusters) versus the estimated values.**

Unfortunately, this is a major problem with evaluating clustering algorithms. Except under highly contrived conditions, we do not know what the true values are for output measures such as the number of clusters. This problem is outlined in ***LXXX-LXXX***. This is why we have developed the concept of the MCC over our last several papers on the subject of OTU clustering.

**A systematic evaluation of the influence of additional variables on performance (i.e., accuracy metrics, RAM, and speed) would provide better insight into expected performance on other data sets. The analyzed data sets do cover a broad range of environments with different types of communities. Even with those data, it would be nice to see performance metrics plotted versus phylogenetic diversity, percentage of sequences in the reference database (for pre-assignment analyses), and sequencing variables (e.g., read length, quality scores). In the simulations, you could additionally introduce realistic sequencing errors evaluate how performance changes.**

As the reviewer correctly points out, there is no ideal dataset. This is why we selected four datasets that represent a variety of environments that are currently popular among microbial ecologists. We also added the simulated datasets (i.e. even and staggered) since these have been used in prior studies evaluating clustering methods. Given that the nature of the algorithm is to optimize the MCC coefficient (or any other coefficient) and we have demonstrated that it is effective at doing this, we don't think it is necessary to add additional datasets to reinfoce the point. To the reviewer's suggestion, we have added the number of OTUs that were found using the OptiClust algorithm and the number of pairwise distances that were equal to and smaller than 0.03 to the summary statistics in Table 1. Unfortunately, some of the datasets are too large and the sequencing data too noisy to permit the construction of the phylogenetic trees that would be needed to calculate the phylogenetic diversity for each study. As for comparing to percentage of sequences in a reference database, we feel that this is would be a misleading point as we have previously shown that reference-based methods are inferior to *de novo* clustering methods (***XXXX***) and the identity of the sequences does not affect the performance of the algorithm (i.e. we could have used synthetic DNA sequences and communities and reached the same conclusion). Finally, the the point regarding read length, quality scores, and sequencing errors we do feel that these analyses are beyond the scope of the current paper. OptiClust only uses pairwise distances as inputs to the clustering and data quality is a very important, but different problem. Given the data provided in Table 1 and Figure 2, we have demonstrated the effect of increasing the number of distances on the performance of the algorithm.

**This may be beyond the scope, but it would be very interesting to see the performance of OptiClust with other criteria as the target of optimization. In other words, how much of the performance is due to targeting MCC as the criterion versus the algorithm itself as an alternative to hierarchical clustering. You could investigate this by comparing the OptiClust algorithm minimizing FPs, for example, as compared to hierarchical clustering minimizing FPs. If this analysis is not performed, I recommend discussing that the algorithm for building clusters and the criterion that is being optimized are evaluated together in your analyses and not disentangled.**

We have added two supplementary figures. The first demonstrates the performance of the algorithm when we optimized clustering based on a variety of other metrics (Figure S1). The second, demonstrates how the MCC values improve when optimizing for the MCC value with each successive iteration (Figure S2).

**Providing the analysis pipeline is great. Users who want to apply the method may like to have more description of the implementation and a stable link to the code (even if this is the same as mothur). To enable applications outside OTU clustering, you should also have a code base that is not embedded in mothur.**

The code for the OptiClust algorithm is available as C++ source code within mothur (e.g. opticluster.cpp at https://github.com/mothur/mothur/tree/master/source). We are reluctant to separate the OptiClust source code from mothur to make it a stand alone tool. This makes maintenance of the software cumbersome and is contrary to the already successful development strategy that we have had for mothur. It should be noted that mothur is wrapped within other tools such as QIIME. Also, mothur can be run from the command line or using scripts. Because of these points, we do not feel it is necessary to separate the code out of the main package.

### Minor comments

**Line 69: Typo: “to be recapitulate”**

This has been fixed.

**Line 132: Not clear why the number of OTUs is mentioned here**

This was meant to indicate that the small difference in MCC values (0.0001) had a relatively small effect on the number of OTUs that result from the algorithm. Although we do not feel that the number of OTUs is a particularly relevant metric of clustering quality, we did feel that it was important to show the effect of small differences in MCC on measures of community richness.

**Line 138: With partial convergence, does the output include the last value for change in MCC?**

Yes, the MCC values are provided as part of the output of running the command.

**Lines 140-150: Mention if agglomerative hierarchical methods are also better than divisive (cite literature or own data).**

We have added a sentence to the Introduction (***LXXX***) indicating that all of the *de novo* methods used by microbial ecologists are agglomerative.

**Lines 151-153: Processing the sequences/OTUs deterministically is a good option for alleviating dependence on data order.**

We have added a clause to this sentence to indicate that this really isn't possible since the algorithms must find a way to break ties when a sequence is equally similar to multiple OTUs.

**Line 161: Typo: “values considerably smaller”**

This has been fixed.

**Line 170: Is disk or CPU also useful to measure?**

In our experience users are limited by RAM and want the methods to complete as quickly as possible. This is shown by the failure of the USEARCH methods to load the datasets in the available RAM or other algorithms to finish in 50 hrs with 48 GB of RAM. The availability of disk space is rarely a problem.

**Lines 182-185: Clarify that this is assuming the pairwise similarities are similar to the analyzed data set. Can you say something about what happens as diversity increases or decreases?**

We have added a line at ***LXXX*** and ***LXXX*** to indicate that speed was proportionate to the number of distances smaller than 0.03. It does not necessarily appear to be related to the number of sequences or the number of OTUs.

**Line 187: Remind the reader that this is “open reference”**

A sentence has been added to this section to clarify the differnces between our method and open reference clustering. Briefly, in open reference clustering sequences are classified and any that do not map to a reference are clustered using a de novo approach. In our approach, all sequences are classified (not mapped) to a broad taxonomic level and then all sequences are clustered using a de novo approach.

**Line 189: Typo: “reduce” should be past tense**

This has been fixed.

**Line 197: Are all of these with taxonomic assignment first? Clarify.**

This has been clarified.

**Line 204: Explain why (i.e., what types of errors the extra clusters represent).**

This has been clarified.

**Line 232: Typo: “effectively ask”**

This has been fixed.

**Line 239: Typo: “are encourage”**

This has been fixed.

## Reviewer 2

### Major comments

**This is a well described, self-enclosed, well implemented piece of work that does not require major revision.  However, there are minor issues which should be addressed.**

### Minor comments

**On page 6 and 7 the authors compare the speed of OptiClust to other methods; however, I do not think Figure 1 really conveys this, mostly due to the log-scaled y-axis for the time panel.  Whilst I recognise why the log-scale has been used, the authors may consider a different way to visually present the data that better conveys the message "The OptiClust algorithm was considerably faster than the hierarchical algorithms and somewhat slower than the heuristic-based algorithms"**

We tried several methods for improving the visualization. First, we made all speed values relative to those in the first column. The problem with this was that the range of values was still considerable and obscured the ability to see the differences. Second, we added grid lines that overlapped with the values from the first column and extended to the right along the plot. This ended up making the figure look overly busy. Third, we attempted to group the methods within the sample types (we had grouped the samples within the methods type), but it was difficult to clearly represent the 13 different methods and 6 samples efficiently. Ultimately, we decided to add a thin gridline at the major y-axis values to give the reader something to anchor their perspective as they looked across the plot to compare the different methods.

**The comparison of RAM used (at the top of page 7) may also be better demonstrated visually.**

At ***LXXX***, our goal wasn't to compare RAM usage between methods. We merely wanted to point out that the USEARCH-based approaches could not handle large datasets without purchasing the 64-bit version of the software and that OTUCLUST and SumClust could not store the data within 45 GB of RAM.

**It wasn't clear to me which software implementation was used for "Nearest Neighbour" and "Furthest Neighbour".**

These were run using the implementations found in mothur (***LXXXX***).

**The authors describe the potential implications of optimising a metric other than MCC - I wonder if they have actually done this?**

We have added text to the manuscript in response to this comment and that of the first reviewer.

### 2017-02-02 submission/Westcott_OptiClust_mSphere_2016.Rmd
Fix cover page formatting
First attempt to respond to initial round of revisions

**Running title:** OptiClust: Optimized Clustering
\vspace{25mm}
Sarah L. Westcott and Patrick D. Schloss${^\dagger}$

@@

A growing number of algorithms have been developed to cluster sequences into OTUs. These algorithms can be classified into three general categories. The first category of algorithms has been termed closed-reference or phylotyping [@Schloss2011; @NavasMolina2013]. Sequences are compared to a reference collection and clustered based on the reference sequences that they are similar to. This approach is fast; however, the method struggles when a sequence is similar to multiple reference sequences that may have different taxonomies and when it is not similar to sequences in the reference. The second category of algorithms has been called *de novo* because they assign sequences to OTUs without the use of a reference [@NavasMolina2013]. These include hierarchical algorithms such as nearest, furthest, and average neighbor [@Schloss2005] and algorithms that employ heuristics such as abundance or distance-based greedy clustering as implemented in USEARCH [@Edgar2010] or VSEARCH [@Rognes2016], Sumaclust, OTUCLUST [@Albanese2015], and Swarm [@Mah2014]. *De novo* methods tend to be more computationally intense and it has proven difficult to know which method generates the best assignments. A third category of algorithm is open-reference clustering, which is a hybrid approach [@NavasMolina2013; @Rideout2014]. Here sequences are assigned to OTUs using closed-reference clustering and sequences that are not within a threshold of a reference sequence are then clustered using a *de novo* approach. This category blends the strengths and weaknesses of the other method and adds the complication that closed-reference and *de novo* clustering use different OTU definitions. These algorithms take different approaches to handling large datasets to minimize the time and memory requirements while attempting to assign sequences to meaningful OTUs.

Several metrics have emerged for assessing the quality of OTU assignment algorithms. These have included the time and memory required to run the algorithm [@Sun2009; @Cai2011; @Rideout2014; @Mah2014], agreement between OTU assignments and the sequences' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016], sensitivity of an algorithm to stochastic processes [@He2015], the number of OTUs generated by the algorithm [@Edgar2013; @Kopylova2016], and the ability to regenerate the assignments made by other algorithms [@Schmidt2014; @Rideout2014]. Unfortunately, these methods fail to directly quantify the quality of the OTU assignments. An algorithm may complete with minimal time and memory requirements or generate an idealized number of OTUs, but the composition of the OTUs could be incorrect. These metrics also tend to be subjective. For instance, a method may appear to be recapitulate the taxonomy of a synthetic community with known taxonomic structure, but do a poor job when applied to real communities with poorly defined taxonomic structure or for sequences that are prone to misclassification. As an alternative, we developed an approach to objectively benchmark the clustering quality of OTU assignments [@Schloss2016a; @Westcott2015; @Schloss2011]. This approach counts the number of true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN) based on the pairwise distances. Sequence pairs that are within the user-specified threshold and are clustered together represent TPs and those in different OTUs are FNs. Those sequence pairs that have a distance larger than the threshold and are not clustered in the same OTU are TNs and those in the same OTU are FPs. These values can be synthesized into a single correlation coefficient, the Matthews' Correlation Coefficient (MCC), which measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix [@Matthews1975]. Consistently, the average neighbor algorithm was identified as among the best or the best algorithm. The distance-based greedy clustering as implemented in VSEARCH has also performed well. The computational resources required to complete the average neighbor algorithm can be significant for large datasets and so there is a need for an algorithm that efficiently produces consistently high quality OTU assignments.

@@

***OptiClust algorithm.***
The OptiClust algorithm uses the pairs of sequences that are within a desired threshold of each other (e.g. 0.03), a list of all sequence names in the dataset, and the metric that should be used to assess clustering quality. A detailed description of the algorithm is provided for a toy dataset in the Supplementary Material. Briefly, the algorithm starts by placing each sequence either within its own OTU or into a single OTU. The algorithm proceeds by interrogating each sequence and re-calculating the metric for the cases where the sequence stays in its current OTU, is moved to each of the other OTUs, or is moved into a new OTU. The location that results in the best clustering quality indicates whether the sequence should remain in its current OTU or be moved to a different or new OTU. Each iteration consists of interrogating every sequence in the dataset. Although numerous options are available within the mothur-based implementation of the algorithm (e.g. sensitivity, specificity, accuracy, F1 score, etc.), the default metric is MCC because it includes all four parameters from the confusion matrix. The algorithm continues until the optimization metric stabilizes or until it reaches a defined stopping criteria.

@@

***OptiClust-generated OTUs are as stable as those from other algorithms.***
One concern that many have with *de novo* clustering algorithms is that their output is sensitive to the initial order of the sequences. An additional concern with the OptiClust algorithm is that it may stabilize at a local optimum. To evaluate these concerns we compared the results obtained using ten randomizations of the order that sequences were given to the algorithm. The median the coefficient of variation across the six datasets for MCC values obtained from the replicate clusterings using OptiClust was `r format(round(stability %>% filter(method=="mcc") %>% select(med_cov_mcc), 1), nsmall=1L)`% (Figure 1). We also measured the coefficient of variation for the number of OTUs across the six datasets for each method. The median coefficient of variation for the number of OTUs generated using OptiClust was `r format(round(stability %>% filter(method=="mcc") %>% select(med_cov_num_otus), 1), nsmall=1L)`%. Confirming our previous results, all of the methods we tested were stable to stochastic processes. Of the methods that involved randomization, the coefficient of variation for MCC values considerably smaller with OptiClust than the other methods and the coefficient of variation for the number of OTUs was comparable to the other methods. The variation observed in clustering quality suggested that the algorithm does not appear to converge to a locally optimum MCC value. More importantly, the random variation does yield output of a similarly high quality.

@@

***Time and memory required to complete Optimization-based clustering scales efficiently.***
Although not as important as the quality of clustering, the amount of time and memory required to assign sequences to OTUs is a legitimate concern. To evaluate how the speed and memory usage scaled with the number of sequences in the dataset, we measured the time required and maximum RAM usage to cluster 20, 40, 60, 80, and 100% of the unique sequences from each of the natural datasets using the OptiClust algorithm (Figure 2). Within each iteration of the algorithm, each sequence is compared to every other sequence and each comparison requires a recalculation of the confusion matrix. This would result in a worst case algorithmic complexity on the order of N^3, where N is the number of unique sequences. Because the algorithm only needs to keep track of the sequence pairs that are within the threshold of each other, it is likely that the implementation of the algorithm is more efficient. To empirically determine the algorithmic complexity, we fit a power law function to the data in Figure 2A. We observed power coefficients between `r paste(format(round(range(speed_coef), 1), nsmall=1L), collapse=" and ")` for the `r paste(names(speed_coef[c(which.min(speed_coef), which.max(speed_coef))]), collapse=" and ")` datasets, respectively. The algorithm requires storing a matrix that contains the pairs of sequences that are close to each other as well as a matrix that indicates which sequences are clustered together. The memory required to store these matrices is on the order of N^2, where N is the number of unique sequences. In fact, when we fit a power law function to the data in Figure 2B, the power coefficients were  `r format(round(median(memory_coef), 1), nsmall=1L)`. This analysis suggests that doubling the number of sequences in a dataset would increase the time required to cluster the data by 4 to 8-fold and increase the RAM required by 4-fold. It is possible that future improvements to the implementation of the algorithm could improve this performance.

@@

***Cluster splitting heuristic generates OTUs that are as good as non-split approach.***
We previously described a heuristic to accelerate OTU assignments where sequences were first classified to taxonomic groups and within each taxon sequences were assigned to OTUs using the average neighbor clustering algorithm [@Schloss2011]. This accelerated the clustering and reduce the memory requirements because the number of unique sequences is effectively reduced by splitting sequences across taxonomic groups. Furthermore, because sequences in different taxonomic groups are assumed to belong to different OTUs they are independent, which permits parallelization and additional reduction in computation time. Reduction in clustering quality are encountered in this approach if there are errors in classification or if two sequences within the desired threshold belong to different taxonomic groups. It is expected that these errors would increase as the taxonomic level goes from kingdom to genus. To characterize the clustering quality, we calculated the MCC values using OptiClust, average neighbor, and DGC with VSEARCH when splitting at each taxonomic level (Figure 3). For each method, the MCC values decreased as the taxonomic resolution increased; however, the decrease in MCC as not as large as the difference between clustering methods. As the resolution of the taxonomic levels increased, the clustering quality remained high, relative to clusters formed from the entire dataset (i.e. kingdom-level). The MCC values when splitting the datasets at the class and genus levels were within `r format(round(min(100*(rel_mcc_data %>% filter(tax_level == 3))$min_mcc), 1), nsmall=1L)` and `r format(round(min(100*(rel_mcc_data %>% filter(tax_level == 6))$min_mcc), 1), nsmall=1L)`%, respectively, of the MCC values obtained from the entire dataset. These decreases in MCC value resulted in the formation of as many as `r format(round(min(100*((rel_otu_data %>% filter(tax_level == 3))$max_num_otus-1), na.rm=T), 1), nsmall=1L)` and `r format(round(min(100*((rel_otu_data %>% filter(tax_level == 6))$max_num_otus-1), na.rm=T), 1), nsmall=1L)`% more OTUs, respectively, than were observed from the entire dataset. For the datasets included in the current analysis, the use of the cluster splitting heuristic was probably not worth the loss in clustering quality. However, as datasets become larger, it may be necessary to use the heuristic to clustering the data into OTUs.

@@

The cluster optimization procedure is dependent on the metric that is chosen for optimization. We employed the MCC because it includes the four values from a confusion matrix. Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled [@Schloss2005; @Schloss2011]. Alternatively, one could optimize based on the sensitivity, specificity, or accuracy, which are each based on two values from the confusion matrix or they could optimize based on the F1 score, which is based on three values from the confusion matrix. Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure. For example, optimizing for sensitivity could lead to a large number of FPs. Since we would like to minimize both FPs and FNs and not just the total number of false assignments, we decided to optimize utilizing the MCC. It is possible that other metrics could be developed and employed for optimization of the clustering.

The OptiClust algorithm is relatively simple. For each sequence it effectively ask whether the MCC value will increase if the sequence is moved to a different OTU including creating a new OTU. If the value does not change, it remains in the current OTU. The algorithm repeats until the MCC value stabilizes. Assuming that the algorithm is seeded with each sequence in a separate OTU, it does not appear that the algorithm converges to a local optimum. Furthermore, execution of the algorithm with different random number generator seeds produces OTU assignments of consistently high quality. Future improvements to the implementation of the algorithm could provide optimization to further improve its speed and susceptibility to find a local optimum. Users are encourage to repeat the OTU assignment several times to confirm that they have found the best OTU assignments.

@@

**Table 1. Description of datasets used to evaluate the OptiClust algorithm and compare its performance to other algorithms.** Each dataset contains sequences from the V4 region of the 16S rRNA gene. The even and staggered datasets were generated by extracting the V4 region from full length reference sequences and the datasets from the natural communities were generated by sequencing the V4 region using a Illumina MiSeq with either paired 150 or 250 nt reads.

@@

data_table <- cbind(dataset_ref, sequencing, data_summary$n_samples,
 										format(data_summary$n_total_seqs, big.mark=','),
 										format(data_summary$n_unique_seqs, big.mark=','))

colnames(data_table) <- c("**Dataset (Ref.)**",
													"**Read Length**",
													"**Samples**",
													"**Total Seqs.**",
													"**Unique Seqs.**")

kable(data_table, row.names=F, align="lcccc")

@@

>>>>

Running title: OptiClust: Optimized Clustering

\vspace{25mm}
Sarah L. Westcott and Patrick D. Schloss${^\dagger}$

@@

A growing number of algorithms have been developed to cluster sequences into OTUs. These algorithms can be classified into three general categories. The first category of algorithms has been termed closed-reference or phylotyping [@Schloss2011; @NavasMolina2013]. Sequences are compared to a reference collection and clustered based on the reference sequences that they are similar to. This approach is fast; however, the method struggles when a sequence is similar to multiple reference sequences that may have different taxonomies and when it is not similar to sequences in the reference. The second category of algorithms has been called *de novo* because they assign sequences to OTUs without the use of a reference [@NavasMolina2013]. These include hierarchical algorithms such as nearest, furthest, and average neighbor [@Schloss2005] and algorithms that employ heuristics such as abundance or distance-based greedy clustering as implemented in USEARCH [@Edgar2010] or VSEARCH [@Rognes2016], Sumaclust, OTUCLUST [@Albanese2015], and Swarm [@Mah2014]. *De novo* methods are agglomerative and tend to be more computationally intense. It has proven difficult to know which method generates the best assignments. A third category of algorithm is open reference clustering, which is a hybrid approach [@NavasMolina2013; @Rideout2014]. Here sequences are assigned to OTUs using closed-reference clustering and sequences that are not within a threshold of a reference sequence are then clustered using a *de novo* approach. This category blends the strengths and weaknesses of the other method and adds the complication that closed-reference and *de novo* clustering use different OTU definitions. These algorithms take different approaches to handling large datasets to minimize the time and memory requirements while attempting to assign sequences to meaningful OTUs.

Several metrics have emerged for assessing the quality of OTU assignment algorithms. These have included the time and memory required to run the algorithm [@Sun2009; @Cai2011; @Rideout2014; @Mah2014], agreement between OTU assignments and the sequences' taxonomy [@Edgar2013; @Mah2014; @Mah2015; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014; @Cai2011; @Sun2011; @White2010; @AlGhalith2016], sensitivity of an algorithm to stochastic processes [@He2015], the number of OTUs generated by the algorithm [@Edgar2013; @Kopylova2016], and the ability to regenerate the assignments made by other algorithms [@Schmidt2014; @Rideout2014]. Unfortunately, these methods fail to directly quantify the quality of the OTU assignments. An algorithm may complete with minimal time and memory requirements or generate an idealized number of OTUs, but the composition of the OTUs could be incorrect. These metrics also tend to be subjective. For instance, a method may appear to recapitulate the taxonomy of a synthetic community with known taxonomic structure, but do a poor job when applied to real communities with poorly defined taxonomic structure or for sequences that are prone to misclassification. As an alternative, we developed an approach to objectively benchmark the clustering quality of OTU assignments [@Schloss2016a; @Westcott2015; @Schloss2011]. This approach counts the number of true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN) based on the pairwise distances. Sequence pairs that are within the user-specified threshold and are clustered together represent TPs and those in different OTUs are FNs. Those sequence pairs that have a distance larger than the threshold and are not clustered in the same OTU are TNs and those in the same OTU are FPs. These values can be synthesized into a single correlation coefficient, the Matthews' Correlation Coefficient (MCC), which measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix [@Matthews1975]. Consistently, the average neighbor algorithm was identified as among the best or the best algorithm. Other hierarchical algorithms such as furthest and nearest neighbor, which seek to minimize the number of FP and FN, respectively, fared significantly worse. The distance-based greedy clustering as implemented in VSEARCH has also performed well. The computational resources required to complete the average neighbor algorithm can be significant for large datasets and so there is a need for an algorithm that efficiently produces consistently high quality OTU assignments.

@@

***OptiClust algorithm.***
The OptiClust algorithm uses the pairs of sequences that are within a desired threshold of each other (e.g. 0.03), a list of all sequence names in the dataset, and the metric that should be used to assess clustering quality. A detailed description of the algorithm is provided for a toy dataset in the Supplementary Material. Briefly, the algorithm starts by placing each sequence either within its own OTU or into a single OTU. The algorithm proceeds by interrogating each sequence and re-calculating the metric for the cases where the sequence stays in its current OTU, is moved to each of the other OTUs, or is moved into a new OTU. The location that results in the best clustering quality indicates whether the sequence should remain in its current OTU or be moved to a different or new OTU. Each iteration consists of interrogating every sequence in the dataset. Although numerous options are available for optimizing the clusters and for assessing the quality of the clusters within the mothur-based implementation of the algorithm (e.g. sensitivity, specificity, accuracy, F1 score, etc.), the default metric for optimization and assessment is MCC because it includes all four parameters from the confusion matrix (Figure S1). The algorithm continues until the optimization metric stabilizes or until it reaches a defined stopping criteria (Figure S2).

@@

***OptiClust-generated OTUs are as stable as those from other algorithms.***
One concern that many have with *de novo* clustering algorithms is that their output is sensitive to the initial order of the sequences because each algorithm must break ties where a sequence could be assigned to multiple OTUs. An additional concern with the OptiClust algorithm is that it may stabilize at a local optimum. To evaluate these concerns we compared the results obtained using ten randomizations of the order that sequences were given to the algorithm. The median the coefficient of variation across the six datasets for MCC values obtained from the replicate clusterings using OptiClust was `r format(round(stability %>% filter(method=="mcc") %>% select(med_cov_mcc), 1), nsmall=1L)`% (Figure 1). We also measured the coefficient of variation for the number of OTUs across the six datasets for each method. The median coefficient of variation for the number of OTUs generated using OptiClust was `r format(round(stability %>% filter(method=="mcc") %>% select(med_cov_num_otus), 1), nsmall=1L)`%. Confirming our previous results, all of the methods we tested were stable to stochastic processes. Of the methods that involved randomization, the coefficient of variation for MCC values was considerably smaller with OptiClust than the other methods and the coefficient of variation for the number of OTUs was comparable to the other methods. The variation observed in clustering quality suggested that the algorithm does not appear to converge to a locally optimum MCC value. More importantly, the random variation does yield output of a similarly high quality.

@@

***Time and memory required to complete Optimization-based clustering scales efficiently.***
Although not as important as the quality of clustering, the amount of time and memory required to assign sequences to OTUs is a legitimate concern. We observed that the time required to complete the OptiClust algorithm (Figure 1C) paralleled the number of pairwise distances that were smaller than 0.03. To further evaluate how the speed and memory usage scaled with the number of sequences in the dataset, we measured the time required and maximum RAM usage to cluster 20, 40, 60, 80, and 100% of the unique sequences from each of the natural datasets using the OptiClust algorithm (Figure 2). Within each iteration of the algorithm, each sequence is compared to every other sequence and each comparison requires a recalculation of the confusion matrix. This would result in a worst case algorithmic complexity on the order of N^3, where N is the number of unique sequences. Because the algorithm only needs to keep track of the sequence pairs that are within the threshold of each other, it is likely that the implementation of the algorithm is more efficient. To empirically determine the algorithmic complexity, we fit a power law function to the data in Figure 2A. We observed power coefficients between `r paste(format(round(range(speed_coef), 1), nsmall=1L), collapse=" and ")` for the `r paste(names(speed_coef[c(which.min(speed_coef), which.max(speed_coef))]), collapse=" and ")` datasets, respectively. The algorithm requires storing a matrix that contains the pairs of sequences that are close to each other as well as a matrix that indicates which sequences are clustered together. The memory required to store these matrices is on the order of N^2, where N is the number of unique sequences. In fact, when we fit a power law function to the data in Figure 2B, the power coefficients were  `r format(round(median(memory_coef), 1), nsmall=1L)`. Using the four natural community datasets, doubling the number of sequences in a dataset would increase the time required to cluster the data by 4 to 8-fold and increase the RAM required by 4-fold. It is possible that future improvements to the implementation of the algorithm could improve this performance.

@@

***Cluster splitting heuristic generates OTUs that are as good as non-split approach.***
We previously described a heuristic to accelerate OTU assignments where sequences were first classified to taxonomic groups and within each taxon sequences were assigned to OTUs using the average neighbor clustering algorithm [@Schloss2011]. This method is similar to open reference clustering except that in our approach all sequences are subjected to *de novo* clustering following classification whereas in open reference clustering only those sequences that cannot be classified are subjected to *de novo* clustering. Our cluster splitting approach accelerated the clustering and reduced the memory requirements because the number of unique sequences is effectively reduced by splitting sequences across taxonomic groups. Furthermore, because sequences in different taxonomic groups are assumed to belong to different OTUs they are independent, which permits parallelization and additional reduction in computation time. Reduction in clustering quality are encountered in this approach if there are errors in classification or if two sequences within the desired threshold belong to different taxonomic groups. It is expected that these errors would increase as the taxonomic level goes from kingdom to genus. To characterize the clustering quality, we classified each sequence at each taxonomic level and calculated the MCC values using OptiClust, average neighbor, and DGC with VSEARCH when splitting at each taxonomic level (Figure 3). For each method, the MCC values decreased as the taxonomic resolution increased; however, the decrease in MCC as not as large as the difference between clustering methods. As the resolution of the taxonomic levels increased, the clustering quality remained high, relative to clusters formed from the entire dataset (i.e. kingdom-level). The MCC values when splitting the datasets at the class and genus levels were within `r format(round(min(100*(rel_mcc_data %>% filter(tax_level == 3))$min_mcc), 1), nsmall=1L)` and `r format(round(min(100*(rel_mcc_data %>% filter(tax_level == 6))$min_mcc), 1), nsmall=1L)`%, respectively, of the MCC values obtained from the entire dataset. These decreases in MCC value resulted in the formation of as many as `r format(round(min(100*((rel_otu_data %>% filter(tax_level == 3))$max_num_otus-1), na.rm=T), 1), nsmall=1L)` and `r format(round(min(100*((rel_otu_data %>% filter(tax_level == 6))$max_num_otus-1), na.rm=T), 1), nsmall=1L)`% more OTUs, respectively, than were observed from the entire dataset. These errors were due to the generation of additional false negatives due to splitting similar sequences into different taxonomic groups. For the datasets included in the current analysis, the use of the cluster splitting heuristic was probably not worth the loss in clustering quality. However, as datasets become larger, it may be necessary to use the heuristic to clustering the data into OTUs.

@@

The cluster optimization procedure is dependent on the metric that is chosen for optimization. We employed the MCC because it includes the four values from a confusion matrix. Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled, respectively [@Schloss2005; @Schloss2011]. Alternatively, one could optimize based on the sensitivity, specificity, or accuracy, which are each based on two values from the confusion matrix or they could optimize based on the F1 score, which is based on three values from the confusion matrix. Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure. For example, optimizing for sensitivity could lead to a large number of FPs. More FPs increases the number of OTUs while more FNs collapses OTUs together. It is difficult to know which is worse since community richness and diversity are linked to the number of OTUs. In addition, increasing the number of FNs would overstate the differences between communities while increasing the number of FPs would overstate their similarity. Therefore, it is important to jointly minimize the number of FPs and FNs. With this in mind, we decided to optimize utilizing the MCC. It is possible that other metrics that balance the four parameters could be developed and employed for optimization of the clustering.

The OptiClust algorithm is relatively simple. For each sequence it effectively asks whether the MCC value will increase if the sequence is moved to a different OTU including creating a new OTU. If the value does not change, it remains in the current OTU. The algorithm repeats until the MCC value stabilizes. Assuming that the algorithm is seeded with each sequence in a separate OTU, it does not appear that the algorithm converges to a local optimum. Furthermore, execution of the algorithm with different random number generator seeds produces OTU assignments of consistently high quality. Future improvements to the implementation of the algorithm could provide optimization to further improve its speed and susceptibility to find a local optimum. Users are encouraged to repeat the OTU assignment several times to confirm that they have found the best OTU assignments.

@@

**Table 1. Description of datasets used to evaluate the OptiClust algorithm and compare its performance to other algorithms.** Each dataset contains sequences from the V4 region of the 16S rRNA gene. The number of distances for each dataset are those that were less than or equal to 0.03. The number of OTUs were determined using the OptiClust algorithm. The even and staggered datasets were generated by extracting the V4 region from full length reference sequences and the datasets from the natural communities were generated by sequencing the V4 region using a Illumina MiSeq with either paired 150 or 250 nt reads.

@@

data_table <- cbind(dataset_ref, sequencing, data_summary$n_samples,
 										format(data_summary$n_total_seqs, big.mark=','),
 										format(data_summary$n_unique_seqs, big.mark=','),
										format(data_summary$n_distances, big.mark=','),
										format(round(data_summary$n_sobs), big.mark=','))

colnames(data_table) <- c("**Dataset (Ref.)**",
													"**Read Length**",
													"**Samples**",
													"**Total Seqs.**",
													"**Unique Seqs.**",
													"**Distances**",
													"**OTUs**")

kable(data_table, row.names=F, align="lcccccc")

@@

\includegraphics[width=6.0in]{../results/figures/performance_other.png}

**Figure S1. The OptiClust algorithm is able to effectively cluster sequences into OTUs by minimizing or maximizing numerous metrics.** Plot of MCC (A), number of OTUs (B), and execution times (C) for the comparison of output from the OptiClust algorithm when to minimizing or maximizing a variety of parameters when applied to four natural and two synthetic datasets. Within mothur, OTU assignments can also be made using other metrics including minimizing false positives and maximizing the specificity, positive predictive value, and true negatives; however, these all resulted in sequences being assigned to separate OTUs, which resulted in no false positives and the maximum number of true negatives. The error bars indicate the range of values observed for 10 replicates.

\includegraphics[width=3.5in]{../results/figures/optimization.png}

**Figure S2. The OptiClust algorithm rapidly converges to optimize the Matthew's Correlation Coefficient.** The six datasets were clustered into OTUs using the OptiClust algorithm seeking to maximize the Matthew's Correlation Coefficient. This was repeated 10 times for each dataset.
